DCMIP2016, Part 1: Models and Equation Sets

### 2016-11-08 DCMIP2016-Part1-Models.tex
- Updates to descriptions of model grids

@@

{\color{red}[ALL] Add a short description of your model grid here.}

@@

@@

The equiangular cubed-sphere grid \citep{sadourny1972conservative, ronchi1996cubed} consists of six Cartesian patches arranged along the faces of a cube which is then inflated onto a spherical shell.  More information on this choice of grid can be found in \cite{ullrich2014global}.  On the equiangular cubed-sphere grid, coordinates are given as $(\alpha, \beta, p)$, with central angles $\alpha, \beta \in [- \frac{\pi}{4}, \frac{\pi}{4}]$ and panel index $i_p \in \{1,2,3,4,5,6\}$.  By convention, we choose panels 1--4 to be along the equator and panels $5$ and $6$ to be centered on the northern and southern pole, respectively.

\subsection{Icosahedral grid}

\subsection{Centroidal Voronoi tessellation grid}

@@

@@

@@

@@

@@

@@

{\color{red}[ULLRICH]}

@@

{\color{red}[RANDALL]}

\subsection{Geophysical Fluid Dynamics Laboratory FV Cubed (GFDL-FV3)}

{\color{red}[HARRIS]}

\subsection{Chombo}

{\color{red}[JOHANSEN]}

@@

\subsection{Global Environmental Multiscale (GEM) Model}

{\color{red}[LEE]}

@@

\subsection{ECMWF PentaRei Finite Volume Model}

{\color{red}[KUEHNLEIN]}

\subsection{Icosahedral Non-hydrostatic (ICON) Model}

{\color{red}[GIORGETTA]}

>>>>

\usepackage{subcaption}

@@

@@

The classic latitude-longitude grid consists of a subdivision of the sphere produced by subdividing along lines of constant latitude and longitude {\color{red}[Figure]}.  Because of the convergence of grid lines near the poles, the operational use of this grid requires that the associated numerical scheme be resilient to arbitrarily small Courant number, or that polar filtering be employed to remove unstable computational modes \citep{lin2004vertically}.  This grid is employed by the UK Met Office \citep{davies2005new}.

@@

The equiangular gnomonic cubed-sphere grid \citep{sadourny1972conservative, ronchi1996cubed} consists of six Cartesian patches arranged along the faces of a cube which is then inflated onto a spherical shell.  More information on this choice of grid can be found in \cite{ullrich2014global}.  On the equiangular cubed-sphere grid, coordinates are given as $(\alpha, \beta, p)$, with central angles $\alpha, \beta \in [- \frac{\pi}{4}, \frac{\pi}{4}]$ and panel index $p$.  The structure of this grid supports refinement through stretching \citep{Schmidt1977,harris2016high} or nesting \citep{harris2013two}.  The Cartesian structure of cubed-sphere grid panels is advantageous for numerical methods that are formulated in Cartesian coordinates, or that utilize dimension splitting.  Nonetheless, special treatment of the panel boundaries is often necessary since they represent coordinate discontinuities. {\color{red}[Cubed-sphere grid figure]}

%FV$^3$ is discretized on the equiangular gnomonic cubed-sphere grid as described in \citet{PL2007JCP}. The gnomonic cubed sphere was found to yield the best balance of uniformity and simplicity for use with the FV algorithm. The structure of this grid supports refinement through stretching \citep{Schmidt1977,harris2016high} or nesting \citep{harris2013two}.  The convention for panel numbering used in FV$^3$ differs from that described above, so that odd-numbered panels are adjacent to the next-higher-numbered panel on their right-hand side, and even-numbered panels are adjacent to the next-higher numbered panel on their top side.

\subsection{Icosahedral (triangular) grid}

For the common global applications, the icosahedral triangular grid is derived from the spherical icosahedron that consists of 20 equilateral spherical triangles, 30 great circle edges and 12 vertices.  These initial triangles are then subdivided repeatedly until the desired mean resolution is obtained. For a single subdivision each edge is divided in $n$ arcs of equal length, thus defining new vertices, which by proper connection to other new vertices result in $n^{2}$ triangles filling the original triangle. By construction the new vertices share 6 triangles, thus the refinement process brakes the initial isotropy of the icosahedron and results in non-equilateral triangles of different sizes.

Several methods are available for subdividing the triangular regions.  One such approach is implemented by the ICON grid generator, which allows an ``arbitrary'' subdivision factor $n$ for the first refinement step only, the so-called root refinement. Typical choices are $n$ = 2, 3 or 5. All additional $m$ refinement steps use $n$ = 2, i.e. are bisection steps. A global grid resulting from a root division factor $n$ and $m$ bisections, denominated as \textit{RnBm} grid,  has $n_c = 20 \cdot n^2 \cdot 2^{2m}$ cells, $n_e = 3/2 \cdot n_c$ edges and $n_v = 10\cdot n^{2}\cdot 2^{2m}+2$ vertices. The anisotropy of global grids is reduced by the spring dynamics of \cite{tomita2001}.  For a discussion of the effective resolution see \cite{dipankar2015large}.  The ICON grid generator further allows for inset regional grids, produced by additional refinement steps that are only applied over a limited region, or set of regions.  The dynamical core then allows for either one-way or two-way coupling of the refined region to the parent model. The current operational numerical weather prediction of the Deutscher Wetterdienst (German Weather Service, DWD) for instance uses a $R3B7$ global grid with 2949120 cells and 13 km mean resolution in combination with a refined region over Europe at 6.5 km resolution.

\subsection{Icosahedral (hexagonal) grid}

\subsection{Constrained Centroidal Voronoi Tessellation (CCVT) grids}

Given a set of $N$ distinct points on the sphere $x_i$ (referred to as the generators, $1 \leq i \leq N$), the \textit{Voronoi tessellation} (or the \textit{Voronoi diagram}) associated with the generators is the set of polygons $\Omega_i$ consisting of all points that are closer (in the sense of great-circle distance) to $x_i$ than any other $x_j$ with $i \neq j$ \citep{okabe2009spatial}.  For a given set of generators, this tiling is unique and completely covers the sphere, and so can be employed in conjunction with many finite volume methods.  However, for an arbitrary set of generators it is easy to produce highly distorted polygons, particularly if the density of generators varies substantially.  This has led to the development of \textit{constrained centroidal Voronoi tessellation (CCVT)} \citep{du2003constrained}, which imposes the additional requirement that the set of generators be coincident with the centroids of each polygon.  Given a desired polygonal density function, several algorithms have been developed to generate CCVTs both in Cartesian and spherical geometry (i.e. for ocean basins or ice sheets) \citep{ringler2008multiresolution}.

\begin{figure}
\begin{center}
\includegraphics[width=2in]{CentroidalVoronoiTessellation.png}
\end{center}
\caption{A constrained centroidal Voronoi tessellation with localized grid density that could be employed in the MPAS model.} \label{fig:CCVT}
\end{figure}

\subsection{Octahedral reduced Gaussian grid}
\label{subsection:octahedralreducedgaussiangrid}

As with the classical reduced Gaussian grid of \cite{hortalsimmonsMWR1991}, the octahedral 
reduced Gaussian grid \citep{malardel2016,smolarkiewiczetalJCP2016} specifies the latitudes according to the 
roots of the Legendre polynomials. 
The two grids differ in the arrangement of the points along the latitudes, which follows a simple
rule for the octahedral grid: starting with 20 points on the first latitude around the poles, four points 
are added with every latitude towards the equator, whereby the spacing between points along the 
latitudes is uniform and there are no points at the equator. The octahedral reduced Gaussian grid 
is suitable for transformations involving spherical harmonics, and has been introduced for global 
medium-range numerical weather prediction with the spectral dynamical core of the IFS at ECMWF in 2016. 
Figure~\ref{fig:octahedralgrid} depicts the octahedral reduced Gaussian grid nodes together with
the edges of the primary mesh as applied in the context of the finite-volume discretisation of FVM (Section~\ref{sec:FVM}).

\begin{figure}
\centering
\begin{subfigure}{.4\textwidth}
  \centering
  \includegraphics[width=5cm]{contributions/FVM/octahedral_gauss_nodes}
 \end{subfigure}%
\begin{subfigure}{.4\textwidth}
  \centering
  \includegraphics[width=4.7cm]{contributions/FVM/octahedral_RGG_white}
 \end{subfigure}
\caption{Locations of the octahedral reduced Gaussian grid nodes (left), and the edges of the primary mesh 
connecting the nodes as applied with the finite-volume discretisation in FVM (right). A coarse grid with only
24 latitudes between pole and equator is used for illustration. The dual mesh resolution of the octahedral
reduced Gaussian grid is about a factor 2 finer at the poles than the equator; see \cite{smolarkiewiczetalJCP2016}.}
\label{fig:octahedralgrid}
\end{figure}

\subsection{Yin-Yang grid} \label{sec:GEM_model_grid}

The overset Yin-Yang grid \citep{kageama2004yinyang} has two Cartesian grid components (subsets of a latitude-longitude grid) which are geometrically identical (see Figure \ref{fig:gem_yinyang}).  These components are combined to cover a spherical surface with partial overlap along their borders.  The Yin component covers the latitude-longitude region
\begin{equation}\label{eq:gem_yinyang}
(-\frac{\pi}{4}-\delta_{\theta} \leq \theta \leq
\frac{\pi}{4}+\delta_{\theta})  \cap
(-\frac{3\pi}{4}-\delta_{\lambda} \leq \lambda \leq
\frac{3\pi}{4}+\delta_{\lambda}),
\end{equation} where $\delta_{\lambda}, \delta_{\theta}$ are small buffers that are proportional to the respective grid-spacings and are required to enforce a minimum overlap in the overset methodology.  For instance, a common configuration employed by the GEM model for DCMIP fixes 
$ \delta_{\theta}=2 $ degrees and $ \delta_{\lambda}=3 \delta_{\theta}$.  The Yang component covers an analogous area, but is rotated so as to cover the region of the sphere outside of the Yin grid.  This grid is employed by the GEM model, utilizing a pair of regional climate models on the two Cartesian patches.

\begin{figure}[]
\includegraphics[height=6cm]{gem_yinyang.png}
\caption{ Yin-Yang grid }
\label{fig:gem_yinyang}
\end{figure}

@@

\subsection{FV$^3$} \label{sec:FV3Equations}

%%%GFDL start
FV$^3$ solves the fully-compressible shallow-atmosphere Euler equations using the vector-invariant equations for horizontal momentum on a mass-based Lagrangian vertical coordinate \citep{L2004MWR}. 
%%%GFDL end

\subsection{FVM} \label{sec:FVMEquations}

The fully compressible Euler equations solved in FVM are given as
\begin{subequations} \label{fvm:compressible}
\begin{align}
 & \pdiff{\mathcal{G}\rho_d}{t} + \nabla \cdot \left(\mathbf{v} \mathcal{G} \rho_d\right) = 0~,
\label{fvm:mass}  \\
& \pdiff{\mathcal{G} \rho_d \mathbf{u}}{t} + \nabla \cdot \left(\mathbf{v} \mathcal{G} \rho_d \mathbf{u} \right) = 
\mathcal{G}\rho_d\left( - \Theta \mathbf{\widetilde{G}} \nabla \phi' - \frac{ \mathbf{g}}{\theta_{a}} \left(\theta'+\theta_a (\epsilon q'_v - q_c - q_r) \right)
- \mathbf{f} \times \left( \mathbf{u} - \frac{\theta}{\theta_a} \mathbf{u}_{a} \right) 
+ \mathbf{M} \right)~,
\label{fvm:momentum} \\
%- \frac{L}{c_p \pi} & \Big( \frac{\Delta q_{vs}}{\Delta t} & + E_r  \Big)
& \pdiff{\mathcal{G}\rho_d \theta'}{t} + \nabla \cdot \left(\mathbf{v}\mathcal{G}\rho_d\,\theta' \right) = 
\mathcal{G}\rho_d\left(-\mathbf{\widetilde{G}}^{T}\mathbf{u} \cdot \nabla \theta_{a}
- \frac{L}{c_p \pi} \left( \frac{\Delta q_{vs}}{\Delta t}  + E_r  \right) %\left(C_d + E_p\right) 
+  \mathcal{H} \right)~,
\label{fvm:thermodynamic} \\
& \phi' = c_{p} \theta_{0} \left[\left( \frac{R_{d}}{p_{0}} \rho_d \theta (1+q_v/\varepsilon) \right)^{R_{d}/c_{v}} - \pi_{a} \right]~,
\label{fvm:gaslaw}
\end{align}
\end{subequations}
which describe the conservation laws of dry mass \eqref{fvm:mass}, momentum \eqref{fvm:momentum},
and dry entropy \eqref{fvm:thermodynamic}.
Dependent variables in \eqref{fvm:compressible} are dry density $\rho_d$, three-dimensional 
physical velocity vector $\mathbf{u}$, potential temperature perturbation $\theta'$, and Exner 
pressure perturbation $\pi'$, 
with the thermodynamic variables related by the gas law 
\eqref{fvm:gaslaw}~\footnote{Note that $\phi'$ represents a normalised Exner pressure perturbation.}.
Mixing ratios of water vapor, cloud condensate, and rain are denoted as $q_v$, $q_c$, and $q_r$,
respectively. All primed variables correspond to deviations from an ambient state (denoted by
subscript "a")  that satisfies 
a balanced subset of \eqref{fvm:compressible}, thus $\psi'=\psi-\psi_a$, where $\psi = u,v,w,\theta,..$~;
see \cite{prusa2008} and \cite{smolarkiewiczJCP2014}. 
The subscript "0" appearing with $\theta_0$ refers to a constant reference value. 
Symbols appearing on the rhs of the momentum equation \eqref{fvm:momentum}  are the
coefficient 
\begin{equation}
\Theta := \frac{\theta\,(1+q_v/\varepsilon)}{\theta_0\,(1+q_t)} %\equiv \frac{\theta_d}{\theta_0}~,
\end{equation}
in front of the pressure gradient term with $\Theta/\theta_0$ the density potential temperature,
the gravity vector $\mathbf{g} \equiv (0,0,-g)$~\footnote{In the shallow- versus deep-atmosphere 
form of the governing equations,  gravity is constant $g \equiv g_c$ or varies with height as 
$g = g_c\,(a/r)^2$, respectively.}, the Coriolis parameter $\mathbf{f}$, and  
$\epsilon=1/\varepsilon-1$ with $\varepsilon = R_d/R_v$. % and $\Upsilon_{C} := {\theta}/{\theta_a}$.
The governing equations \eqref{fvm:compressible} are formulated with respect to a geospherical 
coordinate system and a generalised height-based terrain-following vertical coordinate~\footnote{For
simplicity, the vertical coordinate is assumed to be time-independent in the current presentation.}.
Associated symbols are the Jacobian of the metric tensor $\mathcal{G}$, a matrix of metric coefficients
$\widetilde{\mathbf{G}}$, its transpose $\widetilde{\mathbf{G}}^{T}$, 
and the transformation of the physical to the contravariant velocity 
$\mathbf{v} = \mathbf{\widetilde{G}}^{T}\mathbf{u}$; see \cite{prusa2003} and \cite{kuehnleinJCP2012} for discussion.
%(see \eqref{prusa2003,kuehnleinJCP2012} for discussion). 
The symbol $\mathbf{M}$ in \eqref{fvm:momentum} subsumes metric forces due to the curvature
of the sphere \citep[][]{smolarkiewiczetalJCP2016} and momentum dissipation, whereas $\mathcal{H}$ 
in \eqref{fvm:thermodynamic} represents the diffusion of heat.

%$\mathbf{M}$ and $\mathcal{H}$ subsume various rhs forcings to the
%momentum \eqref{fvm:momentum} and thermodynamic \eqref{fvm:thermodynamic} equations.

\subsection{GEM} \label{sec:GEM_equations}
The equations of GEM model \citep{Girard2014} written out for a log-hydrostatic-pressure like coordinate $\zeta$  system co-rotating with the earth are:
\begin{eqnarray}
\frac{ {d \bf V}_{h} }{dt}+ f {\bf k} \times {\bf{V}_{h}} + R_d T_v\nabla
_\zeta \,  \ln\, p+ \left( 1 + \mu \right) \nabla
_\zeta {\phi} = 0,
\label{eq:gem_eq_1} \\
\frac{d w}{dt}- g \mu = 0,
\label{eq:gem_eq_2} \\
\frac{d}{dt} \ln \left( \pi \pdiff{\ln \pi}{\zeta}\right)
+ \nabla
_{ \zeta }\cdot {\bf V}_{{h}}+\frac{\partial
\dot{ \zeta }}{\partial
\zeta }  = 0,
\label{eq:gem_eq_3} \\
\frac{d \ln T_v}{dt} - \kappa \frac {d \ln p} {dt}=0,
\label{eq:gem_eq_4} \\
R_dT_v+  \frac{p}{\pi} \frac{\partial \phi}{\partial \ln \pi}=0,
\label{eq:gem_eq_5} \\
\frac{d \phi}{dt}-  g w  = 0,
\label{eq:gem_eq_6} \\
1+ \mu - \frac{p}{\pi}   \frac{\partial \ln p}{\partial \ln  \pi}=0,
\label{eq:gem_eq_7} \\
\ln \pi= \zeta+Bs ,
\label{eq:gem_eq_8}
\end{eqnarray}

\noindent with horizontal velocity ${\bf V}_h$, 
generalized vertical velocity $\dot{ \zeta }=\frac{d\zeta}{dt}$, 
vertical velocity $w$,
geopotential $\phi$, 
virtual temperature $T_v$, 
pressure $p$ and
hydrostatic pressure $\pi$. The variable 
$\mu$ is a measure of departure from hydrostatic balance, 
$s$ is related to the surface pressure $\pi_s$ and $B$ is a metric term. The last two variables are defined in (\ref{eq:gem_zeta}).
The equations contain a total time derivative $\frac{ {d } }{dt}$ that expresses the change in time while following the air parcel.

\subsection{ICON} \label{sec:ICONEquations}

The equation system of the ICON model is based upon the prognostic variables suggested by \cite{gassmann2008} 
but uses the two-dimensional rather than the three-dimensional Lamb transformation to convert the nonlinear momentum
advection into a vector-invariant form. So far, the shallow-atmosphere approximation is applied, but there
are plans to add an option for the unapproximated deep-atmosphere equations. The basic equation system reads as follows
\begin{align}\label{1eqmo}
\pdiff{v_n}{t} + \pdiff{K_h}{n} + (\zeta + f)v_t + w\pdiff{v_n}{z} = -c_{pd} \theta_v \pdiff{ \pi}{n} + F(v_n)
\end{align}
\begin{align}\label{3eqmo}
\pdiff{w}{t} + {\bf v_h} \cdot \nabla w + w \pdiff{w}{z} = -c_{pd} \theta_v \pdiff{\pi}{z} - g  
\end{align}
\begin{align}\label{conteq}
\pdiff{\rho}{t} + \nabla \cdot ({\bf v} \rho) = 0
\end{align}
\begin{align}\label{traceq}
\pdiff{\rho q}{t} + \nabla \cdot ({\bf v} \rho q) = S(\rho q)
\end{align}
\begin{align}\label{thdeq1}
\pdiff{\rho \theta_v}{t} + \nabla \cdot ({\bf v} \rho \theta_v) = \tilde{Q}
\end{align}
{As in \cite{gassmann2013},} Eq. (\ref{thdeq1}) is reformulated using the time derivative of
\begin{align}\label{def-exner}
\pi = \left( \frac{R_d}{p_{00}} \rho \theta_v \right)^\frac{R_d}{c_{vd}}
\end{align}
in order to obtain
\begin{align}\label{thdeq}
\pdiff{\pi}{t} + \frac{R_d}{c_{vd}} \frac{\pi}{\rho \theta_v} \nabla \cdot ({\bf v} \rho \theta_v) = \hat{Q} \,,
\end{align}
{which simplifies the implicit numerical treatment of the terms representing vertical sound wave propagation.}
The prognostic variables of the (untransformed) equation system are the horizontal velocity component normal to the triangle edges $v_n$, the
vertical wind component $w$, density $\rho$, and virtual potential temperature $\theta_v$. When coupled with moisture physics,
$\rho$ represents the full air density including liquid and solid hydrometeors. The reconstructed
tangential velocity component is denoted as $v_t$, and in accordance with the model code, it is assumed here that 
$(v_t,v_n,w)$ form a right-handed system. $\bf v_h$ and $\bf v$ denote the horizontal and three-dimensional wind vector,
respectively. The remaining symbols are the vertical vorticity component $\zeta$, 
the Coriolis parameter $f$, the horizontal part of the kinetic energy $K_h = \frac{1}{2} (v_n^2 + v_t^2)$,
the Exner function $\pi$, the specific heat capacities of dry air at constant pressure $c_{pd}$ and volume $c_{vd}$,
the gas constant of dry air $R_d = c_{pd}-c_{vd}$, the gravitational acceleration $g$, {and the reference pressure
$p_{00}$ = 1000 hPa conventionally used for defining the potential temperature.} $\pdiff{}{n}$ denotes a horizontal derivative
in edge-normal direction, i.e. between the two adjacent mass points. Moreover, $F(v_n)$ denotes source
terms for horizontal momentum (to be specified below), and $\tilde{Q}$ and $\hat{Q}$ denote appropriately 
formulated diabatic heat source terms. In the tracer equation $q$ denotes the mixing ration with respect to the full air density $\rho$, and $S(\rho q)$ represents the source term for a tracer $q$.

@@

\subsection{FV3}

Explicit dissipation in FV$^3$ is applied separately to the divergence and to the horizontal fluxes in the governing equations. The D-grid discretization applies no direct implicit dissipation to the divergence, so divergence damping is an intrinsic part of the solver algorithm since otherwise there are no processes by which energy contained in the divergent modes is removed at the grid scale.  FV$^3$ has options for fourth-, sixth-, or eighth-order divergence damping; a second-order option is also available for use in idealized convergence tests, which can be applied in addition to the higher-order diffusion. The monotonicity constraint used in computing the fluxes in the  momentum, thermodynamic, and mass continuity equations is sufficient to damp and stabilize the non-divergent component of the flow. If additional damping is desired, or if the monotonicity constraint has been disabled, there is an option to apply hyperdiffusion to the fluxes in each of these equations, with the exception of the tracer transport, which always uses monotonic transport with no explicit diffusion. This dissipation is of the same order of accuracy as the divergence damping; however the nondimensional coefficient needed is much smaller, usually by at least 3--4 times, than the divergence damping. Both divergence damping and hyperdiffusion are applied along the Lagrangian surfaces and are re-computed every acoustic timestep; there is no explicit damping perpendicular to the Lagrangian surfaces. An option to convert kinetic energy lost to the hyperdiffusion to heat is available. FV$^3$ can also apply a horizontal Smagorinsky eddy diffusion if desired.

A substantial amount of wave absorption is provided by the flexible-lid (constant-pressure) upper boundary;  FV$^3$ also applies second-order diffusion to all fields, except the tracers, to create a sponge layer, typically comprising the top two layers of the domain, to damp other signals reaching the top of the domain. An energy-conserving Rayleigh damping, applied consistently to all three components of the winds, is also available, which is strongest in the top layer of the domain and becomes weaker with distance until reaching a runtime-specified cut-off pressure.

\subsection{GEM}

Via an application of (\ref{eq:scalar_viscosity_direct}),
scalar viscosity is employed for wind components and tracers. 
A vertical sponge layer, also via an application of (\ref{eq:scalar_viscosity_direct}),
is employed on wind components and $T_v$ with a vertical modulation on the 6 top levels 
and a maximum damping coefficient of $.380000 \times 10^{6}\ \mbox{m\textsuperscript{2}/s}$ at the top.
For stabilization purpose, 
the temporal discretization of GEM presented  
in section \ref{sec:GEM_temporal} uses an off-centering parameter $b^{A} = 0.6$.

\subsection{ICON}

The ICON model employs damping and diffusion operators for numerical stabilization and dynamic closure. Details are described in sections 2.4 and 2.5 of \cite{zangl2014}. Here a brief summary:

\emph{Damping: }In the corrector step a fourth-order divergence damping term $F_{d}(\mathbf{v})$ is applied in order to allow calling the computationally more expensive diffusion operator (see below) at the physics time steps only without incurring numerical stability problems under extreme conditions.
\begin{align}
F_{d}(\mathbf{v}) = - f_d \overline{a_c}^2 \nabla \tilde{\nabla} \cdot \left(\nabla \left(\tilde{\nabla} \cdot v + 
\frac{\Delta }{\Delta z} \left(w-\overline{\overline{w_{cc}}^c}^i \right)\right)\right) \,.
\end{align}
$f_d$ typically attains values between $\frac{1}{1000 \Delta t}$ and $\frac{1}{250 \Delta t}$, and $\overline{a_c}$ is the global mean cell area.

Another artificial damping term is Rayleigh damping on $w$ following \cite{klemp2008}, which serves to prevent unphysical reflections of gravity waves at the model top. The Rayleigh damping is restricted to a fixed number of levels below the model top, and the damping coefficient is given by a hyperbolic tangent.

\emph{Diffusion: }The horizontal diffusion consists of a flow dependent second-order Smagorinsky diffusion of velocity ($F_{D2}(v_n)$) and potential temperature  ($F_{D2}(\theta)$) combined with a fourth-order background diffusion of velocity $F_{D4}(v_n)$.
\begin{align} \label{smagn2}
F_{D2}(v_n) = 4 K_h \tilde{\nabla}^2(v_n)
\end{align}
\begin{align}
F_{D2}(\theta) = a_c \tilde{\nabla} \cdot \left(K_h \frac{\Delta \theta}{\Delta n} \right) \,.
\end{align}
\begin{align}
F_{D4}(v_n) = - k_4 a_e^2 \tilde{\nabla}^2 (\tilde{\nabla}^2(v_n)) \,.
\end{align}
An empirically determined offset of $0.75 k_4 a_e$ is subtracted from $K_h$ in order to avoid excessive diffusion under weakly disturbed conditions.

A fourth-order computational diffusion is also available for vertical wind speed $w$. This filter term is needed at resolutions of O(1 km) or finer because the advection of vertical wind speed has no implicit damping of small-scale structures. Consequently diffusion of $w$ is employed for the DCMIP SCELL test.
\begin{align}
F_{D}(w) = - k_w a_c^2 \nabla^2 (\nabla^2(w)) 
\end{align}

@@

In GEM model, shape-preserving solutions are computed for tracers by using the quasi-monotone semi-Lagrangian (QMSL) method \citep{Bermejo1992QMSL}.

@@

FV$^3$ has an option to restore lost energy by the adiabatic dynamics, in whole or a fraction thereof (decided by a namelist option at runtime), by globally adding a Exner-function weighted potential temperature increment. This is only done before the physics is called and is not used in idealized simulations.

@@

\subsection{Semi-implicit time integration of the fully compressible Euler equations in FVM}

In the following we provide an outline of the semi-implicit time stepping scheme for the fully 
compressible Euler equations in FVM (Section~\ref{sec:FVM}). 
A comprehensive discussion of the integration scheme can be found in \cite{smolarkiewiczJCP2014,smolarkiewiczetalJCP2016} 
for dry dynamics; and in \cite{kurowskiJAS2014} and \cite{smolarkiewiczetalMWR2017} for extensions to 
moist dynamics. 
The generic two-time-level second-order template algorithm employed in the integration is given as
\begin{equation}
\psi_{\mbf{i}}^{n+1}= \mathcal{A}_{\mbf{i}}(\widetilde{\psi}^{n},\mathbf{V}^{n+1/2}, (\mathcal{G}\rho_d)^{n}, (\mathcal{G}\rho_d)^{n+1})
+0.5\,\Delta t\,R^{\psi}|_{\mbf{i}}^{\scs n+1}~,
\hspace{1em} \widetilde{\psi}^{n}\,\equiv\,\psi^n+0.5\,\Delta t\,R^{\psi}|^{\scs n}~.
\label{fvm:solscalar}
\end{equation}
In \eqref{fvm:solscalar}, $\psi$ represents the solution variable, $R^{\psi}$ is the respective rhs,
and $\mathcal{A}$ symbolises an advective transport operator assumed here to be the non-oscillatory 
finite-volume MPDATA (Multidimensional Positive Definite Advection Transport Algorithm) scheme 
\citep{smolarkiewiczszmelter2005,kuehnleinJCP2016}~\footnote{Furthermore, 
in \eqref{fvm:solscalar} the vector index $\mathbf{i}$ denotes the spatial position 
on the computational grid, $\Delta t$ is the time step size between levels $n$ and $n+1$.}.

The integration of the system \eqref{fvm:compressible} can basically be divided into three steps. First, the homogenous
mass continuity equation is integrated with $\psi \equiv \rho_d$, $\mathbf{V} \equiv \mathbf{v}\mathcal{G} $,
and $R^{\rho_d} \equiv 0$ in \eqref{fvm:solscalar}. Second, the thermodynamic \eqref{fvm:thermodynamic}, 
momentum \eqref{fvm:momentum}, and moisture equations enter
\eqref{fvm:solscalar} with $\psi = u, v, w, \theta',...$, $\mathbf{V} \equiv \mathbf{v}\mathcal{G}\rho_d$, and the rhs $R^{\psi}$ 
which is generally depending on all prognostic variables. A high degree of implicitness in the representation of the rhs forcings
is achieved by inverting the overall discrete system \eqref{fvm:solscalar} to obtain closed-form expressions for the
velocity updates -- the procedure is facilitated by the co-located arrangement of variables on the computational mesh. %indicated by the spatial mesh vector index..
Retained on the rhs of the derived closed-form velocity expressions is the pressure gradient term. 
The third step in the solution procedure is to formulate an implicit boundary value problem for the pressure variable $\phi'$ using 
an evolutionary form of the equation of state \eqref{fvm:gaslaw}. An $\mathcal{O}(\Delta t^2)$ integration of this 
equation with a Euler backward template algorithm in the spirit of \eqref{fvm:solscalar} leads a Helmholtz equation. 
The associated 3D elliptic boundary value problem is solved iteratively using a bespoke preconditioned Generalised Conjugate 
Residual approach \citep{smolarkiewiczECMWF2004,smolarkiewiczAG2011}.
Nonlinearities in $R^{\psi}$ and the solution-dependent coefficients of the Helmholtz problem are lagged behind and 
executed in an outer iteration.

%The overall integration scheme is 3d implicit with respect to the fast acoustic and buoyant as well as 
%slow rotational modes.

%The integration scheme of FVM approximates the generic conservation law
%\begin{equation}
%\frac{\partial \rho\mathcal{G}\Psi}{\partial t} + \nabla \cdot \left(\rho\mathcal{G}\mathbf{v}  \Psi \right) = \rho\mathcal{G}R^{\Psi}~, %\hspace{1em} 
%\label{scalar}
%\end{equation}
%which represents any of the governing equations \eqref{fvm:mass}-\eqref{fvm:thermodynamic}
%(i.e.~$\Psi = 1 , u, v, w, \theta'$,  as well as corresponding rhs $R^{\Psi}$), by the 
%second-order template algorithm
%\begin{equation}
%\Psi_{\mbf{i}}^{n+1}= \mathcal{A}_{\mbf{i}}(\widetilde{\Psi}^{n},(\rho \mathcal{G} v^{\perp})^{n+1/2}, (\rho\mathcal{G})^{n}, (\rho\mathcal{G})^{n+1})
%+0.5\,\delta t\,R^{\Psi}|

\subsection{Forward-backward vertically-Lagrangian dynamics}

%%%GFDL start

FV$^3$ and its predecessors are integrated using a forward-backwards integration for the Lagrangian dynamics. With the exception of the pressure-gradient force, all of the terms in the momentum, energy, and mass equations are expressable as fluxes, and so can be integrated using the explicit forward-in-time algorithm described by \citet{LR1997QJR}. The horizontal component of the pressure-gradient force is evaluated backwards-in-time using the algorithm of \citet{L1997QJR}; the nonhydrostatic component of the vertical pressure gradient force is evaluated using a semi-implicit solver. This forward-backward timestep is referred to as the ``acoustic'' timestep, although the full solver is advanced on each of these acoustic timesteps. Physics tendencies are applied ``impulsively'' at prescribed intervals, consistent with the forward-in-time discretization; the physics timestep is typically much longer than the acoustic timestep. 
%%%GFDL end

\subsubsection{ICON}

The time stepping scheme of ICON consists of a two-time-level predictor corrector scheme, which is explicit for all terms except for those describing the vertical propagation of sound waves. No time splitting is used with respect to sound waves, because the ration of the maximum wind speed in the mesosphere, which is in part covered by the vertical domain, can be close to one. Instead time splitting is employed to dynamics on the one hand and horizontal diffusion, tracer transport, fast physics on the other hand. Typically a full time step consists of 4 or 5 dynamical sub-steps in which a constant forcing originating from the slow physics is applied. Mass-consistent transport is achieved by passing time-averaged air-mass fluxes from the dynamical sub-steps to the transport scheme.The details of the predictor corrector scheme, including measures to increase the numerical efficiency and to optimize the accuracy, are described in section 2.4 of \cite{zangl2014}.

@@

The CSU model uses an optimized geodesic grid to discretize the sphere, with height as the vertical coordinate. The model is based on the non-hydrostatic Unified System of equations proposed by Arakawa and Konor (2009), which filters vertically propagating sound waves but allows the Lamb wave and does not require a reference state. The horizontal wind field is determined by predicting the vertical component of the vorticity and the divergence of the horizontal wind, and then solving a pair of two-dimensional Poisson equations for a stream function and velocity potential. Time-differencing is based on the third-order Adams-Bashforth  scheme. Horizontal diffusion is included in the form of a $\nabla^4_z()$ operator acting on the vorticity, divergence, potential temperature, and tracer.  
\subsection{Geophysical Fluid Dynamics Laboratory FV Cubed (GFDL-FV$^3$)}

FV$^3$ uses a fully finite-volume discretization of the fully-compressible nonhydrostatic Euler equations on the equiangular gnomonic cubed-sphere grid on a terrain-following Lagrangian vertical coordinate. The flow is entirely along the Lagrangian surfaces, allowing vertical transport to be represented implicitly without additional advection terms. Fluxes are computed using the Piecewise-Parabolic Method of \citet{CW1984JCP} with an optional monotonicity constraint; in nonhydrostatic applications the monotonicity constraint is used primarily for tracer transport. The discretization is on the C-D grid described by \citet{LR1997QJR} which acts as a simplified Riemann solver: the D-grid winds are interpolated to the C-grid and then advanced by half of an acoustic timestep, giving time-centered winds that can then be used to compute the fluxes and advance the flux terms by a full acoustic timestep. Since divergence is effectively invisible to the solver, a divergence damping is applied to control numerical noise as divergent modes cascade to the grid scale.

Implicit viscosity is applied through the monotonicity constraint; if non-monotonic advection is used for the momentum and total air mass a weak explicit hyperviscosity is applied for stability and to alleviate numerical noise. Explicit viscosity is applied every acoustic timestep.

The solver prognoses horizontal winds, in the native Gnomonic local coordinate; virtual potential temperature, which is conserved by the adiabatic dynamics; mass, represented by the difference in hydrostatic pressure between the top and bottom of a grid cell; and tracer mass. The nonhydrostatic solver adds a prognostic vertical velocity and geometric height of each grid cell, which can then be used to compute density. All variables are 3D cell-mean values, except for the horizontal winds, which are 2D face-mean values on their respective staggerings; as a result, vorticity is a 3D cell-mean value.

\subsection{Global Environmental Multiscale (GEM) Model} \label{sec:GEM_core}

In GEM model, we use the Yin-Yang grid and horizontal discretization is done on an Arakawa C grid.
The vertical coordinate $\zeta$ is of a log-hydrostatic-pressure type and vertical discretization is based  
on the Charney-Phillips grid.
A 2 time level semi-Lagrangian implicit time discretization is implemented as described in section \ref{sec:GEM_temporal}. 
A scalar viscosity is employed for wind components and tracers 
via an application of (\ref{eq:scalar_viscosity_direct}). 
Viscosity operations are applied after the completion of the dynamic time step.

@@

@@

@@

\subsection{Finite-volume module of the Integrated Forecasting System}
\label{sec:FVM}

%{\color{red}[KUEHNLEIN]}
